Large and Infinitary Quotient Inductive-Inductive Types

### 2020-05-01 draft/paper.tex
bug
finalization

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

@@

%% ------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage{scalerel}
\usepackage{stmaryrd}
\usepackage{tikz-cd}

\newtheorem{notation}{Notation}

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}
\newcommand{\Inductive}{\mathsf{Inductive}}
\newcommand{\Subtype}{\mathsf{Subtype}}
\newcommand{\subtype}{\mathsf{subtype}}
\newcommand{\NatSig}{\mathsf{NatSig}}
\newcommand{\mi}[1]{\mathit{#1}}

\newcommand{\Sg}{\Sigma}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\lamK}{\mathsf{lam}^{\K}}
\newcommand{\appK}{\mathsf{app}^{\K}}

\newcommand{\arri}{\Rightarrow}
\newcommand{\arre}{\Rightarrow^{\mathsf{ext}}}
\newcommand{\arrinf}{\Rightarrow^{\mathsf{inf}}}
\newcommand{\syn}{\mathsf{syn}}
\newcommand{\SynSig}{\mathsf{SynSig}}

\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}
\newcommand{\bM}{\boldsymbol{\mathsf{M}}}
\newcommand{\bT}{\boldsymbol{\mathsf{T}}}
\newcommand{\bS}{\boldsymbol{\mathsf{S}}}
\newcommand{\bP}{\boldsymbol{\mathsf{P}}}
\newcommand{\bD}{\boldsymbol{\mathsf{D}}}
\newcommand{\bI}{\boldsymbol{\mathsf{I}}}

\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\ulGamma}{\ul{\Gamma}}
\newcommand{\ulDelta}{\ul{\Delta}}
\newcommand{\ulgamma}{\ul{\gamma}}
\newcommand{\ulOmega}{\ul{\Omega}}
\newcommand{\uldelta}{\ul{\delta}}
\newcommand{\ulsigma}{\ul{\sigma}}
\newcommand{\ulnu}{\ul{\nu}}
\newcommand{\ulepsilon}{\ul{\epsilon}}
\newcommand{\ult}{\ul{t}}
\newcommand{\ulu}{\ul{u}}
\newcommand{\ulA}{\ul{A}}
\newcommand{\ula}{\ul{a}}
\newcommand{\ulB}{\ul{B}}
\newcommand{\tos}{\mathsf{tos}}

\newcommand{\coe}{\mathsf{coe}}
\newcommand{\coh}{\mathsf{coh}}
\newcommand{\llb}{\llbracket}
\newcommand{\rrb}{\rrbracket}

%% ------------------------------------------------------------

@@

%% Whereas the initial category can be given by first defining the type of objects
%% (the empty type) and then defining morphisms separately, for an initial model of
%% a type theory given as a CwF, one has to define $\mi{Ob}$, $\mi{Mor}$ and two
%% extra sorts $\mi{Ty}$ and $\mi{Tm}$ \emph{inductive-inductively}, at the same
%% time. This is illustrated by the following excerpt of the signature for CwFs
%% extending the previous signature for categories.
%% \begin{alignat*}{3}
%%   & \mi{Ty}        && : \mi{Ob}\ra \Set \\
%%   & \mi{U}         && : \mi{Ty\,i} \\
%%   & \mi{empty}     && : \mi{Ob} \\
%%   & \mi{ext}       && : (I:\mi{Ob})\ra \mi{Ty}\,I\ra \mi{Ob} \\
%%   & \blank[\blank] && : \mi{Ty}\,J\ra \mi{Mor}\,I\,J\ra \mi{Ty}\,I
%% \end{alignat*}
%% While signatures for type theories usually only contain finitary constructors,
%% some QIITs have infinitary constructors, such as the following signature for
%% countably branching trees where the order of the subtrees does not matter,
%% i.e.\ permuting subtrees yields an equal tree.
%% \begin{alignat*}{3}
%%   & \mi{Tree} && : \U \\
%%   & \mi{leaf}  && : \mi{Tree} \\
%%   & \mi{node} && : (\N\ra \mi{Tree})\ra \mi{Tree} \\
%%   & \mi{perm} && : (g :\N\ra \mi{Tree})(f : \N\ra\N)\ra \mathsf{isIso}\,f\ra \\
%%   & && \hspace{0.8em} \mi{node}\,g = \mi{node}\,(g\circ f)
%% \end{alignat*}
%% Additional examples for infinitary QIITs are real, surreal numbers
%% \cite{HoTTbook}, ordinal numbers \cite{lumsdaineShulman} and the partiality
%% monad \cite{partiality}.

%% \subsection{Contributions}
%% The current paper extends the syntax and semantics of QIITs as previously described
%% in the literature \cite{kaposi2019constructing, qiits, dijkstra2017quotient}. We add the following features:

@@

Additionally, we have an internal type for $\blank\leq\blank$, analogously to
$\blank=\blank$ which is an internalization of definitional equality.
\begin{mathpar}
  \inferrule*{\Gamma \vdash A : \Set_i \\ \Gamma \vdash B : \Set_j}
             {\Gamma \vdash \Subtype\,A\,B : \Set_{\max(i, j)}}

@@

  \inferrule*{\Gamma \vdash t : \Subtype\,A\,B}
             {\Gamma \vdash A \leq B}
\end{mathpar}

@@

Signatures are given as contexts in a certain type theory, the theory of
signatures. We shall abbreviate it as ToS. However, ToS turns out to be a large
infinitary QIIT itself, and we would like to define ToS and a notion of
signature without referring to QIITs, only using features present in cETT. As a
first step, we define a notion of model.

@@

%% \begin{example}{Sort equations}. An interesting example is the term lifting
%% rule for cETT (omitting everything except term lifting):
%% \begin{alignat*}{3}
%%   & cETTSig := \\
%%   & ...\,\ext\,Tm{\Lift} : \Pii\,(\Gamma : Con)\,(\Pii\,(A : Ty{\appitt}\Gamma)((\IdU\,(Tm{\appitt}\Gamma{\appitt}A)\,(Tm{\appitt}\Gamma{\appitt}(\Lift_j{\appitt}A))))
%% \end{alignat*}
%% \end{example}

@@

We make a similar choice in the interpretation of signatures themselves: we use
structured CwFs instead of lex categories. The reason here is that CwFs allow us
to compute induction principles in strictly the same way as one would write in
type theory, since we have $\Ty$ and $\Tm$ for a primitive notion of dependent
objects and morphisms. In contrast, dependent objects in lex categories is a
derived notion, and the induction principles we get are only up to
isomorphism. This issue is perhaps not relevant from a purely categorical
perspective, but we are concerned with eventually implementing QIITs in proof
assistants, so we prefer if our semantics computes strictly.

@@

We define $\bCon : \Set_{\max(i+1,j)+1}$ as $\flCwF_{\max(i+1,j)}$.
%% We define
%% $\flCwF$ as follows.

@@

%% We abbreviate the additional structure on CwFs consisting of $\Sigma$, $\Id$ and
%% $\Kfam$ as fl-structure. We use $\Ty$ in an flCwF to represent ``dependent
%% objects'', and $\Tm$ for ``dependent morphisms''. We use these to define
%% induction.

@@

%% For some $(N,\,z,\,s)$ algebra,
%% $\Ty\,(N,\,z,\,s)$ is defined as $(N^P : N \to \Set)\times N^P\,z \times ((n :
%% N)\to N^P\,n \to N^P(s\,n))$, the type of displayed algebras, and
%% $\Tm\,(N,\,z,\,s)\,(N^P,\,z^P,\,s^P)$ is $(N^S : (n : N)\to N^P\,n) \times (N^S\,z = z^P)
%% \times ((n : N) \to N^S\,(s\,n) = s^P\,(N^S\,n))$.

%% Here, $\Con_{\bGamma}$ is the type of algebras for some signature, e.g.\ the
%% type of natural number algebras in the case of
%% $\mathsf{NatSig}$. $\Ty_{\bGamma}\,\Gamma$ is a displayed algebra over a
%% concrete algebra

%% In the categorical semantics for a concrete signature, this informally means
%% that an algebra supports induction if for any dependent algebra over it (which
%% is a bundle of induction motives and methods), there is a dependent morphism
%% into it (which is a bundle of eliminator functions and their $\beta$-rules).

\begin{theorem}[Equivalence of initiality and induction, c.f. \cite{kaposi2019constructing}]\label{thm:initialind}
An object $\Gamma : \Con_{\bGamma}$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof-irrelevant
predicates. \qed
\end{theorem}

The reason for the ``finite limit CwF'' naming is the following: Clairambault
and Dybjer showed that the 2-category of flCwFs is biequivalent to the
2-category of finitely complete categories
\cite{clairambault2014biequivalence}. In particular, in an flCwF the categorical
product of $\Gamma$ and $\Delta$ can be given as $\Gamma \ext \Kfam\, \Delta$,
and the equalizer of $\sigma$ and $\delta$ as $\Gamma\ext
\Id\,(\K\,\Delta)\,(\lamK\,\sigma)\,(\lamK\,\delta)$.

@@

\begin{definition}[Type categories, c.f.\ \cite{clairambault2014biequivalence}]\label{def:type_categories} For each $\Gamma : \Con$, there is a category
whose objects are types $A : \Ty\,\Gamma$, and morphisms from $A$ to $B$ are
terms $t : \Tm\,(\Gamma\,\ext\,A)\,(B[\p])$. Identity morphisms are given by $\q :
\Tm\,(\Gamma\,\ext\,A)\,(A[\p])$, and composition $t \circ u$ by $t[\p, u]$. The
assignment of type categories to contexts extends to a split indexed category. For
each $\sigma : \Sub\,\Gamma\,\Delta$, there is a functor from $\Ty\,\Delta$ to
$\Ty\,\Gamma$, which sends $A$ to $A[\sigma]$ and $t :
\Tm\,(\Gamma\,\ext\,A)\,(B[\p])$ to $t[\sigma\circ \p, \q]$.
\end{definition}

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types as displayed flCwFs. In ibid.\ substitutions are modeled as strict
morphisms, which we cannot do because our infinitary function types and identity
types cannot be interpreted strictly, as we will see in Section
\ref{sec:infinitaryfunction}. Strict preservation is expressed with metatheoretic equality, which
is respected by all cETT constructions, but isomorphisms are not necessarily
respected. Hence, we need to add additional structure to displayed flCwFs which
expresses preservation of base isomorphisms.

@@

\begin{definition}[Signature for ToS] For each level $j$,
we define $\ToSSig_j : \Sig_{j+i}$, as the signature for the theory of
signatures with external sets in $\Set_j$. This is a large and infinitary QIIT
signature, as we have $\Pie$ and $\Piinf$ abstracting over $A : \Set_j$ and
branching with $A \ra \Ty\,\Gamma$ and $A \ra \Tm\,\Gamma\,\U$ respectively. We
present an excerpt from $\ToSSig_i$ below.
\begin{alignat*}{3}
  & \emptycon && \ext (Con : \U)\\
  &           && \ext (Sub : Con\arri Con \arri \U)\\
  &           && \ext (Ty  : Con\arri\U)\\
  &           && \ext (Tm  : \Pii(\Gamma : Con)(\Ty\,\Gamma \arri \U))\\
  &           && ...\\
  &           && \ext (\Piinf : \Pii(\Gamma : Con)\\
  &           && \hspace{4.2em}(\Pie \Set_j (\lambda\,A.\,(A \arrinf \Ty\,\Gamma)\arri\El(\Ty\,\Gamma))))\\
  &           && ...
\end{alignat*}
\end{definition}

@@

We can convert a signature to a syntactic one by interpreting it in $\syn_j$,
and we can convert in the other direction by using ToS-induction to interpret a
$\Gamma : \Con_{\syn_j}$ in an arbitrary ToS model. This is merely a logical
equivalence, external to cETT (because of unvierse polymorphism), and not an
isomorphism.

@@

Cartmell \cite{gat} defines generalized algebraic theories (GATs) using
type-theoretic syntax. Compared to our QII signatures, his signatures support
sort equations but do not cover infinitary constructors or recursive
equations. A way to encode sort equations in our signatures is using
isomorphisms instead of equalities. In contrast to our algebraic definition,
Cartmell's signatures are given by presyntax, named variables and typing
relations, there is no explicit model theory provided for signatures, and no
explicit term model construction is given. Cartmell focuses instead on showing
that contextual categories serve as classifying categories for GATs.

A more semantic approach to QIITs is given by Altenkirch et
al.\ \cite{qiits}. They generalize the initial algebra semantics of inductive
types to QIITs by considering towers of functors and building complete
categories of algebras from them. Their notion of signature does not enforce
strict positivity, hence describes a larger class of QII signatures. They show
equivalence of initiality and induction, but do not construct initial algebras,
nor would that be possible because of the lack of a positivity restriction.

@@

%% Streicher's initiality proof \cite{streicher2012semantics} is a direct
%% construction of a concrete QIIT: that of the syntax of a type theory. He uses
%% preterms, typing relations and quotients to construct the initial model of type
%% theory in a set-theoretic metalanguage. A recent formalization of the same
%% result for a larger type theory was done by Brunerie et al.\ \cite{brunerie} in
%% Agda with UIP, function extensionality and quotients. Lafont et
%% al.\ \cite{induction-is-enough} construct all finitary inductive-inductive types
%% in Agda using only UIP and function extensionality. Examples of
%% inductive-inductive types are constructed without using UIP by Hugunin
%% \cite{jasper}.

>>>>

\copyrightyear{2020}
\acmYear{2020}
\setcopyright{acmlicensed}
\acmConference[LICS '20]{Proceedings of the 35th Annual ACM/IEEE Symposium on Logic
  in Computer Science (LICS)}{July 8--11, 2020}{Saarbr\"ucken, Germany}
\acmBooktitle{Proceedings of the 35th Annual ACM/IEEE Symposium on Logic in
  Computer Science (LICS '20), July 8--11, 2020, Saarbr\"ucken, Germany}
\acmPrice{15.00}
\acmDOI{10.1145/3373718.3394770}
\acmISBN{978-1-4503-7104-9/20/07}
\startPage{1}

@@

%% --------------------------------------------------------------------------------

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathpartir}
\usepackage{scalerel}
\usepackage{stmaryrd}

%% --------------------------------------------------------------------------------

\newcommand{\refl}{\mathsf{refl}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Ty}{\mathsf{Ty}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\proj}{\mathsf{proj}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\ra}{\rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}
\newcommand{\Inductive}{\mathsf{Inductive}}
\newcommand{\Subtype}{\mathsf{Subtype}}
\newcommand{\subtype}{\mathsf{subtype}}
\newcommand{\NatSig}{\mathsf{NatSig}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\Sg}{\Sigma}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\lamK}{\mathsf{lam}^{\K}}
\newcommand{\appK}{\mathsf{app}^{\K}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\D}{\mathsf{D}}
\renewcommand{\S}{\mathsf{S}}
\newcommand{\arri}{\Rightarrow}
\newcommand{\arre}{\Rightarrow^{\mathsf{ext}}}
\newcommand{\arrinf}{\Rightarrow^{\mathsf{inf}}}
\newcommand{\syn}{\mathsf{syn}}
\newcommand{\SynSig}{\mathsf{SynSig}}
\newcommand{\bCon}{\boldsymbol{\Con}}
\newcommand{\bTy}{\boldsymbol{\Ty}}
\newcommand{\bSub}{\boldsymbol{\Sub}}
\newcommand{\bTm}{\boldsymbol{\Tm}}
\newcommand{\bGamma}{\boldsymbol{\Gamma}}
\newcommand{\bDelta}{\boldsymbol{\Delta}}
\newcommand{\bsigma}{\boldsymbol{\sigma}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bepsilon}{\boldsymbol{\epsilon}}
\newcommand{\bt}{\boldsymbol{t}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bA}{\boldsymbol{A}}
\newcommand{\ba}{\boldsymbol{a}}
\newcommand{\bb}{\boldsymbol{b}}
\newcommand{\bB}{\boldsymbol{B}}
\newcommand{\bid}{\boldsymbol{\id}}
\newcommand{\bemptycon}{\boldsymbol{\emptycon}}
\newcommand{\bSet}{\boldsymbol{\Set}}
\newcommand{\bU}{\boldsymbol{\U}}
\newcommand{\bEl}{\boldsymbol{\El}}
\newcommand{\bPii}{\boldsymbol{\Pi}}
\newcommand{\bPie}{\boldsymbol{\Pie}}
\newcommand{\bPiinf}{\boldsymbol{\Piinf}}
\newcommand{\bappi}{\boldsymbol{\mathsf{app}}}
\newcommand{\blami}{\boldsymbol{\mathsf{lam}}}
\newcommand{\bId}{\boldsymbol{\Id}}
\newcommand{\bM}{\boldsymbol{\mathsf{M}}}
\newcommand{\bT}{\boldsymbol{\mathsf{T}}}
\newcommand{\bS}{\boldsymbol{\mathsf{S}}}
\newcommand{\bP}{\boldsymbol{\mathsf{P}}}
\newcommand{\bD}{\boldsymbol{\mathsf{D}}}
\newcommand{\bI}{\boldsymbol{\mathsf{I}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\ulGamma}{\ul{\Gamma}}
\newcommand{\ulDelta}{\ul{\Delta}}
\newcommand{\ulgamma}{\ul{\gamma}}
\newcommand{\ulOmega}{\ul{\Omega}}
\newcommand{\uldelta}{\ul{\delta}}
\newcommand{\ulsigma}{\ul{\sigma}}
\newcommand{\ulnu}{\ul{\nu}}
\newcommand{\ulepsilon}{\ul{\epsilon}}
\newcommand{\ult}{\ul{t}}
\newcommand{\ulu}{\ul{u}}
\newcommand{\ulA}{\ul{A}}
\newcommand{\ula}{\ul{a}}
\newcommand{\ulB}{\ul{B}}
\newcommand{\tos}{\mathsf{tos}}
\newcommand{\coe}{\mathsf{coe}}
\newcommand{\coh}{\mathsf{coh}}
\newcommand{\llb}{\llbracket}
\newcommand{\rrb}{\rrbracket}

%% --------------------------------------------------------------------------------

@@

@@

Additionally, we have an internal $\Subtype$ type, which internalizes subtyping,
analogously to how $\blank=\blank$ internalizes definitional equality. Hence, we
have analogous reflection and uniqueness rules.
\begin{mathpar}
  \inferrule*{\Gamma \vdash A : \Set_i \\ \Gamma \vdash B : \Set_j}
             {\Gamma \vdash \Subtype\,A\,B : \Set_{\max(i, j)}}

@@

  \inferrule*{\Gamma \vdash t : \Subtype\,A\,B}
             {\Gamma \vdash A \leq B}

  \inferrule*{\Gamma \vdash t : \Subtype\,A\,B \\ \Gamma \vdash u : \Subtype\,A\,B}
             {\Gamma \vdash t \equiv u}
\end{mathpar}

@@

Signatures are given as contexts in a certain type theory, the theory of
signatures. We shall abbreviate it as ToS. However, ToS turns out to be a large
infinitary QIIT itself, and we would like to define ToS and a notion of
signature without referring to QIITs, only using features present in cETT.

In previous works by Cartmell \cite{gat} and Sterling
\cite{sterling2019algebraic}, signatures for generalized algebraic theories are
defined using raw syntax together with well-formedness relations. In this way,
signatures can be specified without already assuming the existence of GATs or
QIITs. However, raw syntax is notoriously difficult to work with, and we prefer
to avoid it altogether.

At this point, we do not actually need \emph{syntactic} signatures, which make
it possible to do induction on signatures. We only need a way to write down
well-formed signatures, and interpret them in arbitrary models of ToS. For this,
a weak Church-like encoding suffices, where a signature is given as a typing
context in an arbitrary model of ToS. For this, we first need to specify the
notion of ToS models. However, this is the \emph{only} piece of information
about ToS which we need to manually provide. Other concepts such as
homomorphisms of ToS models and ToS-induction, will be derived from the
semantics of signatures and self-description in Section \ref{sec:tossig}.

@@

@@

We make a similar choice in the interpretation of signatures themselves: we use
structured CwFs instead of lex categories. The reason here is that CwFs allow us
to compute induction principles in strictly the same way as one would write in
type theory, since we have $\Ty$ and $\Tm$ for a primitive notion of dependent
objects and morphisms. In contrast, dependent objects in lex categories is a
derived notion, and the induction principles we get are only up to
isomorphism. This issue is perhaps not relevant from a purely categorical
perspective, but we are concerned with eventually implementing QIITs in proof
assistants, so we prefer if our semantics computes strictly. This was
demonstrated previously in \cite{hiit}, where we provided a program which
computed types of induction principles from signatures of higher
inductive-inductive types, and we believe that the same could be achieved for
the signatures and semantics described in this paper.

@@

We define $\bCon : \Set_{\max(i+1,j)+1}$ as $\flCwF_{\max(i+1,j)}$.

@@

@@

\begin{theorem}[Equivalence of initiality and induction, c.f. \cite{kaposi2019constructing}]
\label{thm:initialind} An object $\Gamma : \Con_{\bGamma}$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof-irrelevant
predicates. \qed
\end{theorem}

The reason for the ``finite limit CwF'' naming is the following: Clairambault
and Dybjer showed that the 2-category of flCwFs is biequivalent to the
2-category of finitely complete categories
\cite{clairambault2014biequivalence}. In particular, in an flCwF the categorical
product of $\Gamma$ and $\Delta$ can be given as $\Gamma \ext \Kfam\, \Delta$,
and the equalizer of $\sigma$ and $\delta$ as $\Gamma\ext
\Id\,(\K\,\Delta)\,(\lamK\,\sigma)\,(\lamK\,\delta)$. While showing equivalence
of initiality and induction does not need all flCwF components (e.g.\ $\Sigma$
is not needed), we build the full flCwF semantics in order to connect to
Clairambault's and Dybjer's results.

@@

\begin{definition}[Type categories, c.f.\ \cite{clairambault2014biequivalence}]
\label{def:type_categories} For each $\Gamma : \Con$, there is a category
whose objects are types $A : \Ty\,\Gamma$, and morphisms from $A$ to $B$ are
terms $t : \Tm\,(\Gamma\,\ext\,A)\,(B[\p])$. Identity morphisms are given by $\q
: \Tm\,(\Gamma\,\ext\,A)\,(A[\p])$, and composition $t \circ u$ by $t[\p,
  u]$. The assignment of type categories to contexts extends to a split indexed
category. For each $\sigma : \Sub\,\Gamma\,\Delta$, there is a functor from
$\Ty\,\Delta$ to $\Ty\,\Gamma$, which sends $A$ to $A[\sigma]$ and $t :
\Tm\,(\Gamma\,\ext\,A)\,(B[\p])$ to $t[\sigma\circ \p, \q]$.
\end{definition}

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types as displayed flCwFs. In ibid.\ the universe $\U$ in ToS was empty, and all
substitutions were ``neutral'', i.e.\ semantic subsitutions were functors which
may permute, duplicate or forget components of algebras, or freely reinterpret
components, and it is easy to see that all such functors strictly preserve
limits. In contrast, the current $\U$ is not empty: it is closed under identity
and infinitary function types. Hence, substitutions and terms are not neutral
anymore, as they can contain canonical type codes in $\U$. Semantically, these
canonical type codes do not merely reshuffle structure, hence they preserve
limits only weakly. We will return to this in Section
\ref{sec:infinitaryfunction}. We are forced to use a weaker semantics where
fl-structure is not preserved strictly, and we also need to add additional
structure to displayed flCwFs which expresses preservation of base isomorphisms.

@@

\begin{definition}[Signature for ToS] For each level $j$,
we define $\ToSSig_j : \Sig_{j+i}$, as the signature for the theory of
signatures with external sets in $\Set_j$. This is a large and infinitary QIIT
signature, as we have $\Pie$ and $\Piinf$ abstracting over $A : \Set_j$ and
branching with $A \ra \Ty\,\Gamma$ and $A \ra \Tm\,\Gamma\,\U$ respectively. We
present an excerpt from $\ToSSig_j$ below.
\begin{alignat*}{3}
  & \emptycon && \ext (Con : \U)\\
  &           && \ext (Sub : Con\arri Con \arri \U)\\
  &           && \ext (Ty  : Con\arri\U)\\
  &           && \ext (Tm  : \Pii(\Gamma : Con)(\Ty\,\Gamma \arri \U))\\
  &           && ...\\
  &           && \ext (\Piinf : \Pii(\Gamma : Con)\\
  &           && \hspace{4.2em}(\Pie \Set_j (\lambda\,A.\,(A \arrinf \Ty\,\Gamma)\arri\El(\Ty\,\Gamma))))\\
  &           && ...
\end{alignat*}
\end{definition}

@@

We can convert a signature to a syntactic one by interpreting it in $\syn_j$,
and we can convert in the other direction by using ToS-induction to interpret a
$\Gamma : \Con_{\syn_j}$ in an arbitrary ToS model. This is merely a logical
equivalence, external to cETT (because of universe polymorphism), and not an
isomorphism.

@@

Cartmell \cite{gat} defines generalized algebraic theories (GATs) using
type-theoretic syntax. Compared to our QII signatures, he supports infinite
signatures and sort equations but does not cover infinitary constructors or
recursive equations. A way to encode sort equations in our system is using
isomorphisms instead of equalities. In contrast to our algebraic definition,
Cartmell's signatures are given by presyntax, named variables and typing
relations, there is no explicit model theory provided for signatures, and no
explicit term model construction is given. Cartmell focuses instead on showing
that contextual categories serve as classifying categories for GATs.

A more semantic approach to QIITs is given by Altenkirch et
al.\ \cite{qiits}. They generalize the initial algebra semantics of inductive
types to QIITs by considering towers of functors and building complete
categories of algebras from them. Their notion of signature does not enforce
strict positivity, hence describes a larger class of QII signatures. They show
equivalence of initiality and induction, but the lack of a positivity
restriction prevents construction of initial algebras.

@@
